---
title: "Differences in gut microbiota in CD vs nonIBD"
output:
  md_document

bibliography: "../bib/my_library.bib"
---

```{r echo = FALSE, message = FALSE, message = FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 15, fig.height = 15)

library(tidyverse)
library(here)
library(microbiome)
library(vegan)

# import data 
load(here("data/processed/tax_abundances.RDS"))
load(here("data/processed/pathway_abundances.RDS"))
source("https://raw.githubusercontent.com/HenrikEckermann/in_use/master/mb_helper.R")
```


```{r}
source(here("R/data_exploration.R"))
```

# Outcomes from recent study and review: 

1. Gut microbiota composition and functional changes in inflammatory bowel disease and irritable bowel syndrome [@vichvilaGutMicrobiotaComposition2018]

- high resolution shotgun sequencing of 1792 (355 IBD, 412 IBS, 1025 controls)
- of the 102 UC-associated bacterial taxa, 87 were also found to be associated with CD
- 15 UC-specific associations incl. *Bacteroides uniformis* and *Bifidobacterium bifidum*.
- decrease in several butyrate producing bacteria incl. *Faecalibacterium Prausnitzt* (not significant for UC but trend was there in patients with active disease)
- *Bacteroides* species only increased in patients with IBD but not IBS (*Bacteroides fragilis*, *Bacteroides vulgatus*)
- Enterobacteriaceae only increased in CD *Escherichia/Shigella*
- *Bifidobacterium longum* lower in CD
- strain level diversity different (diversity of like pathogenic strain increased vs decreased diversity of beneficial microbes)
- glmnet to differentiate IBS from IBD: mean AUC = 0.91 [0.81 - 0.99] when using microbial compositon. Top20 taxa (highest effect size in prediction models led to similar accuracy (mean AUC 0.9)). adding fecal calprotectin led to highest accuracy.
- 175, 61 or 38 altered pathways, respectively, for CD, UC or IBS



2. Differences in Gut Microbiota in Patients With vs Without Inflammatory Bowel Diseases: a Systematic Review [@pittayanonDifferencesGutMicrobiota2019]

- Phylum: Tenericutes, Lentisphaerae for UC &darr; (not )
- Phylum: Actinobacteria, Firmicutes for CD &darr;
- Phylum: Proteobacteria &uarr; but 5 studies didnt find this and one found decrease in UC
- *Faecalibacterium prausnitzii* 6/11 and 4/10 studies of CD and UC,respectively &darr; (one study for each UC and CD found &uarr; for colonic tissue samples, remaining studies show non-sig &darr;)
- Family: Coriobacteriaceae CD & UC &darr; 
- Family: Christensenellacceae CD &darr;
- Genus: Akkermansia &darr; 
- *Eubacterium rectale* UC &darr;
- *Escherischia Coli* &uarr; 
- Genus: Veillonella CD &uarr;  
- alpha diversity either decreased or not different 
- half studies reported data for both active and inactive IBD whereas the rest reported either active or not active.



# MEDIC DATA 
Keep in mind that I do not have the count data but only relative abundances. So I could not use CLR transformed data for statistical testing or DeSeq2.


## IBD_vs_nonIBD

```{r}
feature_names <- c(names(taxa_by_level), "pathway")
output <- map(feature_names, ~compare_groups(feature_name = .x, task = "IBD_vs_nonIBD"))
```

### Species

#### PCoA (Bray Curtis)

```{r}
output[[1]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[1]][[5]])
output[[1]][[6]]
```


#### Differential abundance plots 

```{r}
output[[1]][[7]]
```


### Genus

#### PCoA (Bray Curtis)

```{r}
output[[2]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[2]][[5]])
output[[2]][[6]]
```


#### Differential abundance plots 

```{r}
output[[2]][[7]]
```


### Family

#### PCoA (Bray Curtis)

```{r}
output[[3]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[3]][[5]])
output[[3]][[6]]
```


#### Differential abundance plots 

```{r}
output[[3]][[7]]
```



### Order

#### PCoA (Bray Curtis)

```{r}
output[[4]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[4]][[5]])
output[[4]][[6]]
```


#### Differential abundance plots 

```{r}
output[[4]][[7]]
```


### Class

#### PCoA (Bray Curtis)

```{r}
output[[5]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[5]][[5]])
output[[5]][[6]]
```


#### Differential abundance plots 

```{r}
output[[5]][[7]]
```

### Phylum

#### PCoA (Bray Curtis)

```{r}
output[[6]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[6]][[5]])
output[[6]][[6]]
```


#### Differential abundance plots 

```{r}
output[[6]][[7]]
```


### Superkingdom

#### PCoA (Bray Curtis)

```{r}
output[[7]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[7]][[5]])
output[[7]][[6]]
```


#### Differential abundance plots 

```{r}
output[[7]][[7]]
```



### Pathway

#### PCoA (Bray Curtis)

```{r}
output[[8]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[8]][[5]])
output[[8]][[6]]
```


#### Differential abundance plots 

```{r}
output[[8]][[7]]
```








## UC_vs_CD



```{r}
feature_names <- c(names(taxa_by_level), "pathway")
output <- map(feature_names, ~compare_groups(feature_name = .x, task = "UC_vs_CD"))
```

### Species

#### PCoA (Bray Curtis)

```{r}
output[[1]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[1]][[5]])
output[[1]][[6]]
```


#### Differential abundance plots 

```{r}
output[[1]][[7]]
```


### Genus

#### PCoA (Bray Curtis)

```{r}
output[[2]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[2]][[5]])
output[[2]][[6]]
```


#### Differential abundance plots 

```{r}
output[[2]][[7]]
```


### Family

#### PCoA (Bray Curtis)

```{r}
output[[3]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[3]][[5]])
output[[3]][[6]]
```


#### Differential abundance plots 

```{r}
output[[3]][[7]]
```



### Order

#### PCoA (Bray Curtis)

```{r}
output[[4]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[4]][[5]])
output[[4]][[6]]
```


#### Differential abundance plots 

```{r}
output[[4]][[7]]
```


### Class

#### PCoA (Bray Curtis)

```{r}
output[[5]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[5]][[5]])
output[[5]][[6]]
```


#### Differential abundance plots 

```{r}
output[[5]][[7]]
```

### Phylum

#### PCoA (Bray Curtis)

```{r}
output[[6]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[6]][[5]])
output[[6]][[6]]
```


#### Differential abundance plots 

```{r}
output[[6]][[7]]
```


### Superkingdom

#### PCoA (Bray Curtis)

```{r}
output[[7]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[7]][[5]])
output[[7]][[6]]
```


#### Differential abundance plots 

```{r}
output[[7]][[7]]
```



### Pathway

#### PCoA (Bray Curtis)

```{r}
output[[8]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[8]][[5]])
output[[8]][[6]]
```


#### Differential abundance plots 

```{r}
output[[8]][[7]]
```










## UC_vs_nonIBD



```{r}
feature_names <- c(names(taxa_by_level), "pathway")
output <- map(feature_names, ~compare_groups(feature_name = .x, task = "UC_vs_nonIBD"))
```

### Species

#### PCoA (Bray Curtis)

```{r}
output[[1]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[1]][[5]])
output[[1]][[6]]
```


#### Differential abundance plots 

```{r}
output[[1]][[7]]
```


### Genus

#### PCoA (Bray Curtis)

```{r}
output[[2]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[2]][[5]])
output[[2]][[6]]
```


#### Differential abundance plots 

```{r}
output[[2]][[7]]
```


### Family

#### PCoA (Bray Curtis)

```{r}
output[[3]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[3]][[5]])
output[[3]][[6]]
```


#### Differential abundance plots 

```{r}
output[[3]][[7]]
```



### Order

#### PCoA (Bray Curtis)

```{r}
output[[4]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[4]][[5]])
output[[4]][[6]]
```


#### Differential abundance plots 

```{r}
output[[4]][[7]]
```


### Class

#### PCoA (Bray Curtis)

```{r}
output[[5]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[5]][[5]])
output[[5]][[6]]
```


#### Differential abundance plots 

```{r}
output[[5]][[7]]
```

### Phylum

#### PCoA (Bray Curtis)

```{r}
output[[6]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[6]][[5]])
output[[6]][[6]]
```


#### Differential abundance plots 

```{r}
output[[6]][[7]]
```


### Superkingdom

#### PCoA (Bray Curtis)

```{r}
output[[7]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[7]][[5]])
output[[7]][[6]]
```


#### Differential abundance plots 

```{r}
output[[7]][[7]]
```



### Pathway

#### PCoA (Bray Curtis)

```{r}
output[[8]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[8]][[5]])
output[[8]][[6]]
```


#### Differential abundance plots 

```{r}
output[[8]][[7]]
```












## CD_vs_nonIBD



```{r}
feature_names <- c(names(taxa_by_level), "pathway")
output <- map(feature_names, ~compare_groups(feature_name = .x, task = "CD_vs_nonIBD"))
```

### Species

#### PCoA (Bray Curtis)

```{r}
output[[1]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[1]][[5]])
output[[1]][[6]]
```


#### Differential abundance plots 

```{r}
output[[1]][[7]]
```


### Genus

#### PCoA (Bray Curtis)

```{r}
output[[2]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[2]][[5]])
output[[2]][[6]]
```


#### Differential abundance plots 

```{r}
output[[2]][[7]]
```


### Family

#### PCoA (Bray Curtis)

```{r}
output[[3]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[3]][[5]])
output[[3]][[6]]
```


#### Differential abundance plots 

```{r}
output[[3]][[7]]
```



### Order

#### PCoA (Bray Curtis)

```{r}
output[[4]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[4]][[5]])
output[[4]][[6]]
```


#### Differential abundance plots 

```{r}
output[[4]][[7]]
```


### Class

#### PCoA (Bray Curtis)

```{r}
output[[5]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[5]][[5]])
output[[5]][[6]]
```


#### Differential abundance plots 

```{r}
output[[5]][[7]]
```

### Phylum

#### PCoA (Bray Curtis)

```{r}
output[[6]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[6]][[5]])
output[[6]][[6]]
```


#### Differential abundance plots 

```{r}
output[[6]][[7]]
```


### Superkingdom

#### PCoA (Bray Curtis)

```{r}
output[[7]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[7]][[5]])
output[[7]][[6]]
```


#### Differential abundance plots 

```{r}
output[[7]][[7]]
```



### Pathway

#### PCoA (Bray Curtis)

```{r}
output[[8]][c(3, 4)]
```



#### PERMANOVA (Bray Curtis)

```{r}
kable(output[[8]][[5]])
output[[8]][[6]]
```


#### Differential abundance plots 

```{r}
output[[8]][[7]]
```

